Concurrent Programming in Harmony

### 2021-12-26 book/paper.tex
Rewrote parts of the Bosco chapter

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses $\mathtt{N}+1$ boolean variables.
Variable <{shared}> is initialized to <{False}> while
<{private}[i]}> for each thread <{i}> is initialized to <{True}>.

@@

\autoref{fig:qtest1} shows a program that
checks concurrent <{put}> operations using
<{N}> threads.  Thread \texttt{putter(<{v}>)}
adds <{v}> to the queue.
Thread \texttt{main()} waits until all \texttt{putter}
threads have finished, and then dequeues <{N}>
items, verifying that they form the set
$\{ 0 .. \mathtt{N} - 1 \}$.

@@

Similarly, \autoref{fig:qtest2} shows a program that
checks concurrent <{get}> operations using
<{N}> threads.
The queue is initialized with the sequence
$0 .. \mathtt{N} - 1$.
Thread \texttt{getter(<{v}>)} removes an entry from the
queue and makes sure that no other thread got the
same entry.
Thread \texttt{main()} waits until all <{getter}>
threads have finished and checks to make sure that
the queue is empty.

@@

\index{logical timestamp}%
We present a protocol preserving these properties that is based on the
work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
In order to tolerate <{F}> failures,
it uses $\mathtt{N} = 2\mathtt{F} + 1$ servers.
In other words, the register survives as long as a strict majority of
its copies survive.
All write operation will be ordered by a unique \emph{logical timestamp}
(see also \autoref{ch:testing}).
Each server maintains not only the value of the object, but also the
timestamp of its corresponding write operation.

@@

In the first phase, a client asks each server for its current
timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So, if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

Suppose client $c'$ is trying to write a value <{v}>.
In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
and <{v}>.  Each server that receives the request compares $(t+1, c')$ against
its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
it adopts the new timestamp and its corresponding value <{v}>.  In either case,
the server responds to the client.
Upon $c'$ receiving a response from $\mathtt{N} - \mathtt{F}$ servers,
the write operation completes.
In case of a read operation, client $c'$ simply \emph{writes back} the
highest timestamp it saw in the first phase along with its corresponding value.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a <{.read}>
request to all servers and then waiting for a response from
<{N - F}> servers.
The <{receive()}> function uses the <{bag.combinations}> method
to find all combinations of subsets of responses of size
$\mathtt{N} - \mathtt{F}$.
The second phase of each operation is similar.

@@

@@

Distributed consensus is the problem of having a collection of processors agree
on a single value over a network.
For example, in state machine replication, the state machines have to agree
on which operation to apply next.
Without failures, this can be solved using leader election: first elect a leader, then have that leader decide a value.
But consensus often has to be done in adverse circumstances,
for example in the face of processor failures.
In this chapter, we will present a simple consensus algorithm that can tolerate
fewer than $1/3^{rd}$ of processors failing by crashing.

More precisely,
constant <{F}> contains the maximum number of failures, and
we will assume there are $<{N}> = 3<{F}> + 1$ processors.
Each processor \emph{proposes} a value, which we assume here to be from the
set \{ 0, 1 \}.
By the usual definition of consensus, we want the following two properties:
\begin{enumerate}
\item \emph{Validity}: a processor can only decide a value that has been proposed;
\item \emph{Agreement}: if two processors decide, then they decide the same value.
\end{enumerate}
The consensus problem is surprisingly tricky to solve in the face of processor
failures and without making assumptions about how long it takes to send
and receive a message.

\autoref{fig:bosco} presents our algorithm.  Besides the network variable, it uses a
shared list of proposals and a shared set of decisions.
To reduce the state space to explore, not all permutations of zeroes and ones are
explored.  With 5 processors ($\mathtt{F} = 2)$, say, we only explore the cases where
no processors propose 1, where exactly one processors proposes 1, and where 2 processors
proposes 1.
In this particular algorithm, all messages are broadcast to all processors, so they do
not require a destination address.
The invariants we want to maintain
on those variables is that $<{decisions}> \subseteq <{proposals}>$
and $|<{decisions}>| \leq 1$.  Since the initial value of <{decisions}> is
$\emptyset$, the invariants clearly hold initially.

The $\mathtt{N}$ processors go through a sequence of \emph{rounds} in which they wait for
$<{N}> - <{F}>$ messages, update their state based on the messages, and
broadcast messages containing their new state.
The reason that a processor waits for $<{N}> - <{F}>$ rather than <{N}>
messages is because of failures: up to $<{F}>$ processors may never send a message
and so it would be unwise to wait for all $<{N}>$.  You might be tempted to use
a timer and time out on waiting for a particular processor.  But how would you initialize
that timer?  While we will assume that the network is reliable, there is no guarantee
that messages arrive within a particular time.
We call a set of $<{N}> - <{F}>$ processors a \emph{quorum}.
A quorum must suffice for the algorithm to make progress.

The state of a processor consists of its current round number (initially 0)
and an estimate (initially the proposal).  Therefore, messages contain
a round number and an estimate. To start things, each processor first
broadcasts its initial round number and initial estimate.
The number of rounds that are necessary to achieve consensus is not bounded.  But
Harmony can only check finite models, so there is a constant \texttt{NROUNDS} that
limits the number of rounds.  It may be that no decisions are made, but that does not
violate either Validity or Agreement.  We only check that \emph{if} decisions are
made, they satisfy Validity and Agreement.

In Line~22, a processor waits for <{N - F}> messages using
the Harmony <{atomically when exists}> statement.
Since Harmony has to check all possible executions of the protocol, the
<{receive(round, k)}> method returns all \emph{subbags} of messages
for the given round that have size <{k = N - F}>.  The method uses a
dictionary comprehension to filter out all messages for the given <{round}>
and then uses the <{bag.combinations}> method to find all combinations of size <{k}>.
The <{atomically when exists}> statement waits until there is at least one such combination and
then chooses an element, which is bound to the <{quorum}> variable.
The body of the statement is then executed atomically.
This is usually how distributed algorithms are modeled, because they can only interact
through the network.  There is no need to interleave the different processes other
than when messages are delivered.  By executing the body atomically, a lot of
unnecessary interleavings are avoided and this reduces the state space that must
be explored by the model checker significantly.

The body of the <{atomically when exists}> statement contains the core of the algorithm.
Note that <{N - F = 2F + 1}>, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

@@

While one can run this code within little time for $\mathtt{F} = 2$, for
$\mathtt{F} = 3$ the state space to explore is already quite large.
One way to reduce the state space to explore is the following realization:
each processor only considers messages for the round that it is in.
If a message is for an old round, the processor will ignore it;
if a message is for a future round, the processor will buffer it.
So, one can simplify the model and have each processor wait
for \emph{all} <{N}> messages in a round
instead of $\mathtt{N} - \mathtt{F}$.
It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
out of those <{N}> messages, but executions in which some processors
are left behind in all rounds are no longer considered.
It still includes executions where some subset of $\mathtt{N} - \mathtt{F}$
processors only choose each other messages and essentially ignore the
messages of the remaining <{F}> processors, so the resulting model
is just as good.

\autoref{fig:bosco2} shows the code for this model.  Running this with
$\mathtt{F} = 3$ does not take very long and this approach is a good
blueprint for testing other round-based protocols (of which there are
many).

\section*{Exercises}
\begin{problems}
\item The algorithm as given works in the face of crash failures.
A more challenging class to tolerate are \emph{arbitrary failures} in
which up to <{F}> processors may send arbitrary messages, including
conflicting messages to different peers (equivocation).
The algorithm can tolerate those failures if you use $\mathtt{N} = 5\mathtt{F} - 1$ processors instead of $\mathtt{N} = 3\mathtt{F} - 1$.  Check that.
\item In 1983, Michael Ben-Or presented a randomized algorithm that can
tolerate crash failures with
just $\mathtt{N} = 2\mathtt{F} - 1$ processors~\cite{BenOr83}.
Implement this algorithm.
\end{problems}

@@

Paxos~\cite{Paxos} is the most well-known family of consensus protocols for
environments in which few or no assumptions are made about timing.
In this chapter we present a basic version of a Paxos protocol, one that is
\emph{single-decree} (only tries to make a single decision).
It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
In order to tolerate <{F}> crash failures, you need at least $<{F}>+1$ leaders
and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so
in total only $2<{F}> + 1$ independently failing processors are needed.
Here we provide only a rudimentary introduction to Paxos; for more detailed
information refer to~\cite{Paxos}.

>>>>

\autoref{fig:tas} goes on to implement mutual exclusion for
a set of <{N}> threads.
The approach is called \emph{spinlock},
\index{spinlock}%
because each thread is ``spinning'' (executing a tight loop) until
it can acquire the lock.
The program uses $<{N}>+1$ boolean variables.
Variable <{shared}> is initialized to <{False}> while
<{private}[i]}> for each thread <{i}> is initialized to <{True}>.

@@

\autoref{fig:qtest1} shows a program that
checks concurrent <{put}> operations using
<{N}> threads.  Thread \texttt{putter(<{v}>)}
adds <{v}> to the queue.
Thread \texttt{main()} waits until all \texttt{putter}
threads have finished, and then dequeues <{N}>
items, verifying that they form the set
$\{ 0 .. <{N}> - 1 \}$.

@@

Similarly, \autoref{fig:qtest2} shows a program that
checks concurrent <{get}> operations using
<{N}> threads.
The queue is initialized with the sequence
$0 .. <{N}> - 1$.
Thread \texttt{getter(<{v}>)} removes an entry from the
queue and makes sure that no other thread got the
same entry.
Thread \texttt{main()} waits until all <{getter}>
threads have finished and checks to make sure that
the queue is empty.

@@

\index{logical timestamp}%
We present a protocol preserving these properties that is based on the
work by Hagit Attiya, Amotz Bar-Noy, and Danny Dolev~\cite{ABD95}.
In order to tolerate <{F}> failures,
it uses $<{N}> = 2<{F}> + 1$ servers.
In other words, the register survives as long as a strict majority of
its copies survive.
All write operation will be ordered by a unique \emph{logical timestamp}
(see also \autoref{ch:testing}).
Each server maintains not only the value of the object, but also the
timestamp of its corresponding write operation.

@@

In the first phase, a client asks each server for its current
timestamp and value.  After receiving $<{N}> - <{F}>$ responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So, if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

Suppose client $c'$ is trying to write a value <{v}>.
In phase 2, client $c'$ broadcasts a request containing timestamp $(t+1, c')$
and <{v}>.  Each server that receives the request compares $(t+1, c')$ against
its current timestamp.  If $(t+1, c')$ is larger than its current timestamp,
it adopts the new timestamp and its corresponding value <{v}>.  In either case,
the server responds to the client.
Upon $c'$ receiving a response from $<{N}> - <{F}>$ servers,
the write operation completes.
In case of a read operation, client $c'$ simply \emph{writes back} the
highest timestamp it saw in the first phase along with its corresponding value.

@@

Read and write operations are both invoked with a unique identifier
<{uid}>.  Both start by broadcasting a <{.read}>
request to all servers and then waiting for a response from
<{N - F}> servers.
The <{receive()}> function uses the <{bag.combinations}> method
to find all combinations of subsets of responses of size
$<{N}> - <{F}>$.
The second phase of each operation is similar.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/consensus.png}
\end{center}
\begin{code}
<{:input:../code/consensus.hny}>
\end{code}
\caption{\harmonylink{code/consensus.hny} Distributed consensus code and DFA}
\label{fig:consensus}
\end{figure}

Distributed consensus is the problem of having a collection of processors agree
on a single value over a network.
For example, in state machine replication, the state machines have to agree
on which operation to apply next.
Without failures, this can be solved using leader election: first elect a leader, then have that leader decide a value.
But consensus often has to be done in adverse circumstances,
for example in the face of processor failures.

Each processor \emph{proposes} a value, which we assume here to be from the
set \{ 0, 1 \}.
By the usual definition of consensus, we want the following three properties:
\begin{enumerate}
\item \emph{Validity}: a processor can only decide a value that has been proposed;
\item \emph{Agreement}: if two processors decide, then they decide the same value.
\item \emph{Termination}: each processor eventually decides.
\end{enumerate}
The consensus problem is impossible to solve in the face of processor
failures and without making assumptions about how long it takes to send
and receive a message~\cite{FLP85}.
Here we will not worry about \emph{Termination}.

\autoref{fig:consensus} presents a specification for binary
consensus---the proposals are from the set <{{0, 2}}>
In this case there are four processors.  The proposal of processor <{i}>
is in <{proposals[i]}>.
The <{decision}> is chosen from the set of proposals.
Each processor may or may not print the decision---capturing the
absence of the \emph{Termination} property.
It may be that no decisions are made, but that does not
violate either Validity or Agreement.
Thus the behavior of the program is to first print the array of
proposals, followed by some subset of processors printing their decision.
Notice the following properties:
\begin{itemize}
\item there are $16 = 2^4$ possible proposal configurations;
\item all processors that decide decide the same value;
\item if all processors propose 0, then all processors that decide decide 0;
\item if all processors propose 1, then all processors that decide decide 1.
\end{itemize}

This is just the specification---in practice we do not have a shared
variable in which we can store the decision a priori.
We will present a simple consensus algorithm that can tolerate
fewer than $1/3^{rd}$ of processors failing by crashing.
More precisely,
constant <{F}> contains the maximum number of failures, and
we will assume there are $<{N}> = 3<{F}> + 1$ processors.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/bosco.png}
\end{center}
\caption{The DFA for \autoref{fig:bosco}}
\label{fig:boscodfa}
\end{figure}

\autoref{fig:bosco} presents our algorithm.  Besides the <{network}>
variable,
it uses a shared list of proposals and a shared set of decisions.
In this particular algorithm, all messages are broadcast to all processors, so they do
not require a destination address.
The $<{N}>$ processors go through a sequence of \emph{rounds} in which they wait for
<{N - F}> messages, update their state based on the messages, and
broadcast messages containing their new state.
The reason that a processor waits for <{N - F}> rather than <{N}>
messages is because of failures: up to <{F}> processors may never send a message
and so it would be unwise to wait for all <{N}>.  You might be tempted to use
a timer and time out on waiting for a particular processor.  But how would you initialize
that timer?  While we will assume that the network is reliable, there is no guarantee
that messages arrive within a particular time.
We call a set of <{N - F}> processors a \emph{quorum}.
A quorum must suffice for the algorithm to make progress.

The state of a processor consists of its current round number (initially 0)
and an estimate (initially the proposal).  Therefore, messages contain
a round number and an estimate. To start things, each processor first
broadcasts its initial round number and initial estimate.
The number of rounds that are necessary to achieve consensus is not bounded.  But
Harmony can only check finite models, so there is a constant \texttt{NROUNDS} that
limits the number of rounds.

In Line~21, a processor waits for <{N - F}> messages using
the Harmony <{atomically when exists}> statement.
Since Harmony has to check all possible executions of the protocol, the
<{receive(round, k)}> method returns all \emph{subbags} of messages
for the given round that have size <{k = N - F}>.  The method uses a
dictionary comprehension to filter out all messages for the given <{round}>
and then uses the <{bag.combinations}> method to find all combinations of size <{k}>.
The <{atomically when exists}> statement waits until there is at least one such combination and
then chooses an element, which is bound to the <{quorum}> variable.
The body of the statement is then executed atomically.
This is usually how distributed algorithms are modeled, because they can only interact
through the network.  There is no need to interleave the different processes other
than when messages are delivered.  By executing the body atomically, a lot of
unnecessary interleavings are avoided and this reduces the state space that must
be explored by the model checker significantly.

The body of the <{atomically when exists}> statement contains the core of the algorithm.
Note that $<{N}> - <{F}> = 2<{F}> + 1$, so that the number of
messages is guaranteed to be odd.  Also, because there are only 0 and 1 values, there
must exist a majority of zeroes or ones.  Variable <{count[0]}> stores the number
of zeroes and <{count[1]}> stores the number of ones received in the round.
The rules of the algorithm are simple:
\begin{itemize}
\item update <{estimate}> to be the majority value;
\item if the quorum is unanimous, decide the value.
\end{itemize}
After that, proceed with the next round.

To check for correct behavior, run the following two commands:
\begin{code}
\begin{verbatim}
$ harmony -o consensus.hfa code/consensus.hny
$ harmony -B consensus.hfa code/bosco.hny
\end{verbatim}
\end{code}

You will note that the second command prints a warning:
``\texttt{behavior warning: strict subset of specified behavior}.''
Thus, the set of behaviors that our algorithm generates is a subset
of the behavior that the specification allows.
\autoref{fig:boscodfa} shows the behavior, and indeed it is not the
same as the behavior of \autoref{fig:consensus}.
This is because in our algorithm the outcome is decided a priori if more than
twothirds of the processors have the same proposal, whereas in the
consensus specification the outcome is only decided a priori if the
processors are initially unanimous.

@@

While one can run this code within little time for <{F = 1}>, for
<{F = 2}> the state space to explore is already quite large.
One way to reduce the state space to explore is the following realization:
each processor only considers messages for the round that it is in.
If a message is for an old round, the processor will ignore it;
if a message is for a future round, the processor will buffer it.
So, one can simplify the model and have each processor wait
for \emph{all} <{N}> messages in a round
instead of <{N - F}>.
It would still have to choose to consider just <{N - F}>
out of those <{N}> messages, but executions in which some processors
are left behind in all rounds are no longer considered.
It still includes executions where some subset of <{N - F}>
processors only choose each other messages and essentially ignore the
messages of the remaining <{F}> processors, so the resulting model
is just as good.

Another way to reduce the state space to explore is to leverage
symmetry.  First of all, it does not matter who proposes a particular
value.  Also, the values 0 and 1 are not important to how the
protocol operates.  So, with 5 processors (<{F = 2}>), say, we only
need to explore the cases where no processors propose 1, where
exactly one processors proposes 1, and where 2 processors proposes 1.

\autoref{fig:bosco2} shows the code for this optimized model.
Running this with <{F = 2}> does not take very long and this approach
is a good blueprint for testing other round-based protocols (of
which there are many).

\section*{Exercises}
\begin{problems}
\item The algorithm as given works in the face of crash failures.
A more challenging class to tolerate are \emph{arbitrary failures} in
which up to <{F}> processors may send arbitrary messages, including
conflicting messages to different peers (equivocation).
The algorithm can tolerate those failures if you use $<{N}> = 5<{F}> - 1$ processors instead of $<{N}> = 3<{F}> - 1$.  Check that.
\item In 1983, Michael Ben-Or presented a randomized algorithm that can
tolerate crash failures with
just $<{N}> = 2<{F}> - 1$ processors~\cite{BenOr83}.
Implement this algorithm.
\end{problems}

@@

Paxos~\cite{Paxos} is the most well-known family of consensus protocols for
environments in which few or no assumptions are made about timing.
In this chapter we present a basic version of a Paxos protocol, one that is
\emph{single-decree} (only tries to make a single decision).
It uses two kinds of processors: \emph{leaders} and \emph{acceptors}.
In order to tolerate <{F}> crash failures, you need at least $<{F}> + 1$ leaders
and $2<{F}> + 1$ acceptors, but leaders and acceptors can be colocated, so
in total only $2<{F}> + 1$ independently failing processors are needed.
Here we provide only a rudimentary introduction to Paxos; for more detailed
information refer to~\cite{Paxos}.
